---
title: "chat_openai"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chat_openai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ellmer)
```

## Chat with an LLM

### 1. Initialise the session

To start chatting with an LLM, use the one of the chat functions from `ellmer`
- The model must be supported by the service you have authenticate to
- You can set a system prompt that will define the overall behaviour / response
  of the LLM. This is not the chat input!
```{r}

chat <- chat_azure_openai(
  model = "gpt-5-mini",
  system_prompt = "Be terse"
)
```

### 2. Send your first prompt

You can now ask the LLM a question or give it a task
```{r}
answer <- chat$chat("Which are the 3 largest states in the USA by area?")
```

### 3. Continue the conversation

The `chat` object remembers the previous conversation. This mean your next input does not need to provide the previous context.
```{r}
answer <- chat$chat("Can you re-rank these three by population")
```

This history is kept until the chat session is re-initialised. This happens when
  - You overwrite or `chat` the variable in Step 1 
  - You create a new chat object and use that one instead
  - You restart R

## Creating custom tools

You can provide the LLM access to custom R functions by sharing information about 
what function does (input and output). The LLM can request to run them based on 
the prompt, and use the results for to help shape the answer. To be clear, 
the LLM is not doing any function calculations, R is doing this on your computer.

### 1. Define an R function
```{r}
# Define a regular function to use as a tool (something an LLM would struggle with)
lucky_numbers <- function(n = 3) {
  if (n < 1) {
    return("You must request at least 1 number")
  }
  set.seed(Sys.Date())
  sample(1:100, n)
}

```

#### 2. Convert it into a tool
Use the `tool` function to create the tool from the regular function 

- The description should be clear as this is what the LLM will use to decide
whether it wants to use it or not
- The arguments will be provided as text by the LLM based on the prompt
  - Provide a named list using the function arguments you defined
  - Use any of the available type converters so the LLM can generate an
  appropriate input from the description you provide inside
```{r}
# Create the tool
lucky_numbers_tool <- tool(
  lucky_numbers,
  name = "lucky_numbers",
  description = "Returns a list of n lucky numbers (changes daily)",
  arguments = list(
    n = type_integer("The number of lucky numbers to return. Defaults to 3")
  )
)
```
For a full list of types, check out https://ellmer.tidyverse.org/articles/tool-calling.html#tool-inputs-and-outputs

#### 3. Register the tool for use
```{r}
chat$register_tool(lucky_numbers)
```

#### 4. Chat with the LLM
Based on the prompt, the LLM can now decide to use your tools, which you will 
see popping up in the console if activated during a chat session.
```{r}
chat$chat("What are my lucky numbers?")
```


