---
title: "LLM intro template - Solution"
author: "PJ"
date: "2026-03-09"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,  comment = "#>")
library(edAI)
```

## Intro

This worksheet will guide you through the basics of working with LLMs from 
within R. Given you are using the custom `edAI` library, some parts of this
process have been simplified.

## 1 - Connecting to models hosted on Azure
These exercises assume you have been given access to an Azure instance that is
hosting an LLM. You will **need an API key** in order to proceed

Start by running the `edAI_setup()`function and provide the following info
- username: you can pick whichever username you like, but know this will be
used to track your token usage
- api key: the key provided to you by the instructors 

```{r}
edAI_setup()
```


## 2 - Simple chat with an LLM

### 2.1 - Initialise the session
Run the `edAI_init()` function from the `edAI` package to authenticate with
the Azure cloud

```{r}
edAI_init()
```
### 2.2 - Connect to a chat model

To start chatting with an LLM, use the one of the chat functions from `ellmer`
- Load the `ellmer` library
- Use the `chat_azure_openai()` function to establish a connection and save the
  connection to a variable called `conn`
- Request access to the "gpt-5-mini" model

```{r}
library(ellmer)
conn <- chat_azure_openai(model = "gpt-5-mini")
```

### 2.3. - Send your first prompt

- Using the `conn` object you just created, access the `chat` function it contains
using the `$` operator
- Add a prompt that asks which nerve enervates the biceps brachii muscle 
- Store the result inside a new variable called `answer1`

```{r}
answer1 <- conn$chat("Which nerve enervates the biceps brachii muscle")
```

### 2.3 - Continue the conversation

The `conn` object remembers the conversation which means the next prompt
can build off the last answer without having to provide more context.
- Ask the model which other muscles are enervated by this nerve
- Store the answer in a variable called `answer2`
```{r}
answer2 <- conn$chat("Which other muscles are enervated by this nerve")
```

This history is kept until the chat session is re-initialised. This happens when
  - You overwrite or `conn` the variable you created in the first step
  - You create a new chat object and use that one instead
  - You restart R
  
## 3 - Parsing JSON and CSV

So far the AI has returned plain text, which needs a human to read and process
it before code can continue with the output. Let's for the output to become
structured. 

Before we use the AI, let's first practice parsing JSON and CSV strings to R
objects.

### 3.1 - Parsing CSV
- You normally read CSV files from disk by providing a path to the file, but you 
  can also read them from a string of text using the `textConnection()` function
- Your task is to write a single line of text inside the `string` variable that
  will be converted into a data frame with the follow properties
  - column names: 'id' and 'value'
  - First row of data: id = 1, value = 5
  - Second row of data: id = 2, value = 7
- Check the new variable called `result_table` to see if it worked

Tip: Instead of using real line breaks, insert the line break character `\n`

The resulting table should look like this
| id| val|
|--:|---:|
|  1|   5|
|  2|   7|

```{r}
string <- "id,val\n1,5\n2,7"
result_table <- read.csv(textConnection(string)) 
print(result_table)
```

### 3.2 - Parsing JSON
- Reading JSON from a string is simple. All we need is the `fromJSON()` function 
  from the `jsonlite` library
- Your task is to write a single line of text inside the `string` variable that
  will be converted into an R list with the follow items
  - attempt: 1
  - letters: A, B and C
- Check the new variable called `result_table` to see if it worked

Tips

- JSON ignore line breaks, so you don't need to use `\n`
- If you need to use quotes (e.g. `"`) in a string, you need to
  - Escape using `\` --> "this quote \" is part of the text"
  - Use single quotes for the string and double for the text 
    --> 'this quote " is part of the text'
    
The resulting list should look like this
$attempt
[1] 1

$letters
[1] "A" "B" "C"

```{r}
library(jsonlite)
string <- '{"attempt": 1,"letters":["A","B","C"]}'
result_list <- fromJSON(string)
print(result_list)
```

## 4 - Let the LLM return stuctured data and parse it

Combining all previous steps, we'll now prompt the LLM, make sure the output 
is structured, and then parse it as an R object.
 
- Create a new Azure LLM object called `symptoms` which can take prompts
- The prompt will ask to return a list of up to 5 symptoms for a given disease 
  (e.g. cardiac infarction)
- Structure the prompt such that it will return text that can be parsed as a CSV
  file which, when read into a data frame will have two columns:
  - symptom: the name of the symptom
  - occurrence: score between 1 - 4 with 1 being nearly always present and 4 rarely present
- When finished, you should have a variable called `symptoms_table` that contains
  and R data frame with the result
  
The resulting table should look something like this
                              symptom occurrence
1              Chest pain or pressure         1
2       Shortness of breath (dyspnea)         2
3              Diaphoresis (sweating)         2
4 Radiating pain to left arm/jaw/neck         3
5 Syncope or profound lightheadedness         4

Tip: LLMs know how CSV files are structured, so you don't need to provide an
example, but you do need to describe the expected content (columns and values)

```{r}
symptoms <- chat_azure_openai(model = "gpt-5-mini")
llm_output <- symptoms$chat("Give me a list of up to 5 symptoms for cardiac infarction. For each
rate the occurrence between 1 - 4 with 1 being nearly always present and 4 rarely present . 
Provide the output as a CSV file that can be parsed with the following columns: symptom,occurence")
symptoms_table <- read.csv(textConnection(llm_output))
print(symptoms_table)
```


