---
title: "Accessing a Web API with httr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{edAI_api}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(httr2)
```

## 1 - What is a web API

A web API (Application Programming Interface) is a way for computers to request
and share data across the internet. Many websites that curate data and allow you
to search / download it, also provide and API to automate this process through
code, bypassing the need for manual interaction via the website.

An API call consist of two stages

- **request**: This is the info you send to a server to request specific data
- **response**: This is the (meta)data returned by the server in response to a
  request

There are a few things you need to send a basic web API request

- **The base URL**: This is the link that points to the API server. This URL is
  often different from the web-interface URL
- **Endpoint**: This is the resource you want to access. APIs typically have
  multiple endpoints / resources
- **Query parameters**: Specific filters or other details needed to retrieve the
  data of interest only

Some APIs also require

- **Authentication details**: Many APIs require you to authenticate, just as you
  would login to a website to access resources
- **body**: A set of additional data (often structured as JSON) to be sent along
  with the request
- _This list is not exhaustive, but will not be explored further here_

## 2 - Using httr2 to call an API

The `httr2` library in R provides all functions you'll need to fetch data from a
web API.

Let's start by building a simple request to a public API (no authentication
needed). You should open the documentation website to see what the API expects
as input and returns as output.

### 2.1 - Request

- Source: The Human Phenotype Ontology
- Documentation: https://clinicaltables.nlm.nih.gov/apidoc/hpo/v3/doc.html
- Base URL: https://clinicaltables.nlm.nih.gov/api/hpo/v3/search

```{r}
req_object <- request("https://clinicaltables.nlm.nih.gov/api/hpo/v3/search") |> 
  req_url_query(terms = "fainting")

print(req_object)
```

- `request`: takes the base URL
- `req_url_query`: a list of keys (parameters) and values. In this case the
  documentation specified that `terms` is a required parameter

Printing this object will show you the full URL that contains all combined info.

Note that this simple API only has a **single endpoint** `search`, which means
the base URL is just that endpoint. If there were **multiple endpoints**, we can
extend the base URL using the `req_url_path_append` function.

```{r}
req_object <- request("https://clinicaltables.nlm.nih.gov/api/hpo/v3") |> 
  req_url_path_append("search") |> 
  req_url_query(terms = "fainting")

print(req_object)
```

_The result is identical to the previous step_

### 2.2 - Response

Most modern APIs return their results as JSON files, though it is important to
check this in the documentation.

If the request was successful, we can extract the data from the response

```{r}
result <- req_perform(req_object)
result <- resp_body_json(result)

print(result)
```

- `req_perform()`: This executes the request and fetches the response data
- `resp_body_json`: Extracts JSON data from the response and turns it into an R
  list

### 2.3 - Try it yourself

Modify the request such that it will

- search for terms that contain the text `head`
- Only return the first 5 results (check the documentation)

Tip: You can chain all commands (request and response) together using `|>`

```{r}
result <- request("https://clinicaltables.nlm.nih.gov/api/hpo/v3/search") |> 
  req_url_query(terms = "head", maxList = 5) |> 
  req_perform() |> resp_body_json()

print(result)
```

## 3 - Fetch data from the Clinical Trials API

You will explore the ClinicalTrials.gov API to fetch results on clinical trials
registered with this organisation.

This API contains much more data, and is organised into multiple endpoints
(resources) so you will have to extend the base URL using the
`req_url_path_append` function.

Start by opening the
[documentation](https://clinicaltrials.gov/data-api/api#extapi)

- The base URL for this api is: https://clinicaltrials.gov/api/v2/

### 3.1 - Return the version information from the API

Call the _version_ endpoint. Tip: You don't need any query parameters.

```{r}
request("https://clinicaltrials.gov/api/v2/") |> 
  req_url_path_append("version") |> 
  req_perform() |> resp_body_json()
```

### 3.2 - Return results for a specific study

Using the _studies_ endpoint. Get the results for the study with ID NCT04846335.

Note that you are not going to search, but access a _nested_ endpoint, so you
will have to supply multiple arguments to `req_url_path_append`

```{r}
result <- request("https://clinicaltrials.gov/api/v2/") |> 
  req_url_path_append("studies", "NCT04846335") |> 
  req_perform() |> resp_body_json()

print(result)
```

### 3.3 - Search for studies on Progeria

Using the _studies_ endpoint, search for the following

- Condition: Progeria syndrome
- Only return completed studies

You will need to use both the endpoint and query functions and should expand the
documentation for the study endpoint to find details on names and values of
query parameters.

```{r}
result <- request("https://clinicaltrials.gov/api/v2/") |> 
  req_url_path_append("studies") |> 
  req_url_query(query.cond = "Progeria", filter.overallStatus = "COMPLETED") |> 
  req_perform() |> resp_body_json()

print(result)
```

## 4 - Recreating rentrez library functions

The `rentrez` library in R allows you to interact with the entrez API from NCBI.
However, these functions are simply wrappers around basic API commands we can
recreate with `httr2`

### 4.1 - Search for a MeSH term ID

- Navigate to the
  [Entrez ESearch documentation](https://www.ncbi.nlm.nih.gov/books/NBK25499/#_chapter4_ESearch_)
- Use the following criteria
  - Query the `mesh` database
  - Search for "long covid"
  - Make sure to return JSON (tip: check optional parameters)
  - Sort the result by relevance
- Extract the top ID from the returned result

```{r}
result <- request("https://eutils.ncbi.nlm.nih.gov/entrez/eutils/") |> 
  req_url_path_append("esearch") |> 
  req_url_query(
    db = "mesh", 
    term = "long covid", 
    retmode = "json", 
    sort = "relevance") |> 
  req_perform() |> resp_body_json()

mesh_id <- result$esearchresult$idlist[[1]]
```

### 4.2 - Get MeSH term ID details

Use the ID from the previous step and switch to the `summary` endpoint to return
MeSH info for this ID

```{r}
result <- request("https://eutils.ncbi.nlm.nih.gov/entrez/eutils/") |> 
  req_url_path_append("esummary") |> 
  req_url_query(db = "mesh", id = mesh_id, retmode = "json") |> 
  req_perform() |> resp_body_json()

print(result)
```
