---
title: "edAI_llm_agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{edAI_llm_agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, eval = FALSE)
```

```{r setup}
library(edAI)
library(ellmer)
library(glue)
library(jsonlite)
edAI_init()
```

An AI agent can run and complete a complex task (semi) autonomously by using a
combination of AI models, logic, tools, memory or state.Building an agent from
scratch is a big undertaking, so here we'll explore the very basics using a
simple scenario.

## Scenario - Medical case variant generator

Imagine building a pipeline that takes in a case of a specific disease
presentation and tries to predict the diagnosis. There are some hallmark disease
characteristics, but there could be many others individual characteristics that
can vary between patients with the disease and don't change the base diagnosis,
though they can affect other aspects like severity, prognosis, phenotype etc.

We want to build an AI agent that can generate test cases for this pipeline that
all should lead to the same diagnosis, but varies over individual presentation
characteristics.

## Tools

This dummy example only has one tool available `ai_parse_json`. This will take a
JSON string, checks if it's valid and if not tries to fix any syntax errors. In
case it does not work after several ties or the string is not JSON at all, there
won't be an error but a failure is registered in the return.

_Note that although this tools uses AI, it's not an agent because it has no real
autonomy, the decisions are hard-coded and only completes a single, simple task_

```{r}
ai_parse_json <- function(string, tries = 3) {
  result <- NULL
  for (i in 1:tries) {
    # Try parsing without AI first
    if (jsonlite::validate(string)) {
      result <- fromJSON(string)
    }

    # Fix JSON with AI if previous step failed
    if (is.null(result)) {
      fixJSON <- chat_azure_openai(
        system_prompt = "Try and fix the provided JSON syntax errors and return valid JSON without anything else that could prevent immediate parsing. Do not try and repair a string that is clearly not JSON or would require adding / deleting data, in that case return nothing!",
        model = "gpt-5-mini"
      )
      string <- fixJSON$chat(string)
      if (string == "") {
        # AI thinks it's not a JSON file, stop trying
        break()
      }
    } else {
      # Stop trying if parsing worked
      break()
    }
  }

  return(list(success = !is.null(result), string = string, result = result))
}
```

## AI Agent

Below is the code that creates a very basic agent that completes the goal of
building the cases.

There are logical steps that are currently hard-coded but could be taken over by
a reasoning LLM as well to improve the quality and risk of issues.

### Input

These are the variables provided as input, either directly by a user, code or
another agent.

```{r}
disease <- "osteomyelitis"
number_of_var <- 5
n_cases <- 10
```

### Process

```{r}
# Use an LLM to generate patient characteristics that vary among patients with the same diagnosis
var_prompt <- glue(
  "Consider {disease} and generate a list of {number_of_var} patient characteristics, symptoms or variables that NEW patients can present with but can differ between all kinds of patients with this disease. Changing these should not alter the diagnosis, but could alter progression, outcome, onset etc. Limit the variables to the history taking part or physical examination. You will be asked to provide examples, option or data ranges. Use medically valid values. Report using the provided JSON template"
)
conn <- chat_azure_openai(
  system_prompt = glue(
    'Respond using the JSON template below to return info on {number_of_var} variables but don\'t add anything outside JSON! Replace each "var" with the name of a characteristic (use snake case) and pick the correct type (text, categorical, integer or float) and parameters.\n\nTEMPLATE\n{{"var":{{"type":"text","examples":["example1","example2"]}},"var":{{"type":"categorical","options":["option1","option1"]}},"var":{{"type":"integer","min":0,"max":100},"var":{{"type":"float","min":5.1,"max":7.1}}}}'
  ),
  model = "gpt-5-mini"
)
var_response <- conn$chat(var_prompt)

# Parse the JSON and fix if needed using the tool
var_list <- ai_parse_json(var_response)

if (!var_list$success) {
  stop("Agent failed at JSON parsing step")
}

# Use the set of variables / characteristics to generate a template scenario
template_prompt <- paste(
  'You are given a list of common patient characteristics that can vary between people diagnosed with',
  disease,
  '. Write a general NEW case presentation template (no other info, just presentation) that incorporates all of these characteristics using placeholders that allows us to create many valid variations of the same case later. Do not include the diagnosis.\n\nCharacteristics to use:',
  var_list$string,
  '\n\nTo Return\nOnly return the template, nothing else. Placeholders need to be put between curly brackets. Example: "A {gender} patient of {age} years old ...". ',
  collapse = "\n"
)

conn2 <- chat_azure_openai(
  system_prompt = 'Make sure the language reflects professional medical writing and verify that all placeholders are formatted like {characteristic}. Only use the provided characteristic names, don\'t add any new ones. Adjust the length of the case template the to number of placeholders to keep it short',
  model = "gpt-5-mini"
)

template <- conn2$chat(template_prompt)

cases <- sapply(1:n_cases, function(x) {
  glue_vars <- lapply(var_list$result, function(var) {
    if (var$type %in% c("text", "categorical")) {
      sample(var[[2]], 1)
    } else if (var$type == "integer") {
      sample(var$min:var$max, 1)
    } else {
      runif(1, var$min, var$max) |> round(1)
    }
  })

  glue_data(glue_vars, template)
})

result <- list(
  disease = disease,
  var_list = var_list,
  template = template,
  example_cases = cases
)
```

### Output

```{r}
cat("TEMPLATE\n")
result$template
cat("\nEXAMPLE CASE\n")
result$example_cases[5] |> cat()
```

## Next steps

The agent above is still very simple and does not check all components of
modern, complex agents. For example, there is no autonomous decision-making yet
or AI driven self correction.

There are many ways you could improve this agents capabilities. Think of
something that would and see if you can tweak the code.
