---
title: "edAI_llm_rag"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{edAI_llm_rag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, eval = FALSE)
library(edAI)
library(ragnar)
library(ellmer)
edAI_init()
```

## 1 - Overview

When creating LLM prompts, it is important to provide enough context to get an
accurate answer. However, sometimes the context we need is part of an existing
document, website, ... , and we want to provide this instead. This works for
small amounts of text, but if the context becomes too large, LLMs can't process
this all. We can solve this issue by extracting only relevant pieces from our
sources before sending it to the LLM. This process is called **Retrieval
Augmented Generation** or RAG for short.

We will be using wrapping functions from the `edAI` library which in turn call
functions from the [ragnar](https://ragnar.tidyverse.org/) library. Once you get
more familiar with RAG, you can use the `ragnar` library directly.

### 1.1 - High-level Steps

**Preparation**

1. Extract all text from a source or convert into a text format
2. Divide the text into chunks and LLM can manage
3. Generate a numeric vector for each text chunk using an embedding model
4. Store the vectors + text inside a store (e.g. database)
5. Index the store for fast retrieval

**Retrieval**

1. Convert the prompt to an embedding vector using the same embedding model
2. Search the store for similar vectors
3. Return the text associated with best matching vectors
4. Append the context to the prompt

## 2 - Vector store

The vector store will hold all text and associated embedding vectors. Many
implementations are possible, but most often a database is used. In simple cases
we can use file-based database like `duckdb` or `sqlite`.

### 2.1 - Create a duckdb vector store

Let's use the `edAI_rag_store` function from the `edAI` library to create a
duckdb vector store.

```{r}
rag_store <- edAI_rag_store("demo_store.duckdb")
```

- Provide the path to a new or existing store
- The function expects you to have access to an Azure OpenAI embedding model

### 2.2 - Add text / documents to the vector store

Let's extract all text from a Wikipedia page on the topic of appendectomy. We'll
use the `edAI_wiki_page()` function (which calls the Wikipedia API using
`httr2`)

```{r}
wiki_page <- edAI_wiki_page("https://en.wikipedia.org/wiki/Appendectomy")
cat(wiki_page)
```

Let's now add this to our vector store using the `edAI_rag_insert()` function

- **store**: A vector store object generated by `edAI_rag_store`
- **new_data**: Either one more strings of text, or a link to a file on disk
  (text file, markdown, PDF, ...)
- **origin**: If you directly provide text, you need to set the origin. This
  will ensure the database knows which source / documents the data originated
  from. You can skip this if you provide a file (in which case the file name is
  used)

```{r}
edAI_rag_insert(store = rag_store, new_data = wiki_page, origin = "appendectomy")
```

- This wrapper function will take case of text chunking and embedding and will
  insert the data into the database and index the results
- Depending on the size of the text, this can take a while

### 2.3 - Retrieve text chunks from a vector store

Now we have a vector store with files in it, we can search it for relevant
chunks. For this we'll use the `ragnar_retrieve` function from the `ragnar`
library directly.

- **store**: The vector store to search
- **prompt**: The prompt you plan on sending to the LLM

For demo purposes, we'll set `deoverlap = FALSE`. By default, `ragnar` will
paste together consecutive chunks

```{r}
prompt <- "What type of incision is best for an appendectomy?"
relevant_chunks <- ragnar_retrieve(
    store = rag_store, 
    text = prompt, 
    deoverlap = FALSE
    )
```

### 2.4 - Evaluate quality of returned matches

Let's explore the top match, and its cosine distance, which is an indication of
how well the match is. Cosine Similarity / Distance are values between 0 - 1.

The table below shows a general guidance for score interpretation, but quality
ca depend on the store content and prompt used. It is therefore recommended to
create calibration prompts where you check thinks you know are or are not in the
store and use those scores as a reference.

| Cosine Similarity | Cosine Distance | Meaning                        |
| ----------------- | --------------- | ------------------------------ |
| 1.0               | 0.0             | Identical vectors (best match) |
| 0.8               | 0.2             | Very similar                   |
| 0.5               | 0.5             | Moderately similar             |
| 0.0               | 1.0             | Unrelated                      |

```{r}
cat(relevant_chunks$text[1])
relevant_chunks$cosine_distance[1]
```

_Both the text and the cosine distance indicate a strong match_

#### 2.5 - Update the prompt and re-evaluate the match / score

- Ask about cholecystectomy incisions
- Ask what the ingredient for a haggis are

## 3 - Add RAG as a tool to an LLM search

Once you have a vector store, you can add RAG as a tool to an `ellmer` chat
session. You need to explain to the LLM when to use the store. In the example
below we'll ask to always search the store first.

```{r}
conn <- chat_azure_openai(
  system_prompt = "You are a helpful assistant. Before answering any question, always use the retrieve tool to search the knowledge store for relevant information. Base your answer on the retrieved content and be terse",
  model = "gpt-5-mini"
)

ragnar_register_tool_retrieve(
    conn, 
    edAI_rag_store("demo_store.duckdb"), 
    store_description = "the knowledge store"
)

answer <- conn$chat("Can you remove an appendix in a pregnant woman?")
```

## 4 - Exercise

A RAG store can contain many different documents and there is no requirement
that they are related as the aim of the search is to only return relevant
chunks.

### 4.1 Create a new store

Create a new duckdb vector store called `test_store.duckdb` and save the store
info to a variable

```{r}
test_store <- edAI_rag_store("test_store.duckdb")
```

### 4.2 - Add data to the store

Add the following to the vector store

- One Wikipedia page on a medical topic of your choice
- One Wikipedia page unrelated to medicine
- A custom file called `unknown_info.txt` which you create and to which you add
  a very unique sentence or two that are extremely unlikely to be found on the
  internet

```{r}
edAI_rag_insert(
  store = test_store, 
  new_data = edAI_wiki_page("https://en.wikipedia.org/wiki/Fontan_procedure"), 
  origin = "wiki_fontan"
)
edAI_rag_insert(
  store = test_store, 
  new_data = edAI_wiki_page("https://en.wikipedia.org/wiki/Cobblestone"), 
  origin = "wiki_cobblestone"
)

edAI_rag_insert(
  store = test_store, 
  new_data = prompt_file <- system.file("rmarkdown", "templates", "edAI_llm_rag", "skeleton", "unknown_info.txt", package = "edAI")
)
```

### 4.3 - Search for relevant chunks

Write a few prompts that would match the content of one or more of the documents
you added to the database. Explore the following items in the result object

- origin
- cosine_distance
- text

```{r}
prompt <- "What are the advantages of cobblestone roads over other road types"
relevant_chunks <- ragnar_retrieve(store = test_store,text = prompt)
```

## 4.4 - Couple an LLM to your store

Register you store with an LLM and then ask it a question that can only be
answered if you now the info contained in your `unknown_info.txt` file.

```{r}
conn <- chat_azure_openai(
  system_prompt = "You are a helpful assistant. Before answering any question, always use the retrieve tool to search the knowledge store for relevant information. Base your answer on the retrieved content and be terse",
  model = "gpt-5-mini"
)

ragnar_register_tool_retrieve(
    conn, 
    test_store, 
    store_description = "the knowledge store"
)

answer <- conn$chat("What's the name of the person with the golden hat and pink bow tie")
```
