---
title: "LLM_extra"
---


```{r setup, include=FALSE}
library(ellmer)
library(glue)
library(jsonlite)
library(rentrez)
library(xml2)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)
edAI_init()
```

## 1 - Separate prompt files

Writing long pieces of text inside a quoted R string can be hard and might 
clutter you coding script. Therefore, it's easier to create a separate text 
file, write your prompts as normal, and then read it in.

### 1.1 - Simple file

- Create a new text file called `prompt_simple.txt` in the same folder where you
saved this file. You can use the Files pane in RStudio if you want.
- Inside this file, add a very simple prompt.
- Use the `readLines` function to now read in your file, and save it in a variable
called `prompt`
- Use the `paste` function to _collapse_ all lines into a single string with
new line separators `\n` in between and save this again to the `prompt` variable 
- Use the `cat` function to print the text with line breaks to the console
```{r}
# Replace with your own file if you like
prompt_file <- system.file("rmarkdown", "templates", "edai_llm_extra", "skeleton", "prompt_simple.txt", package = "edAI")

prompt <- readLines(prompt_file) 
prompt <- paste(prompt, collapse = "\n")
cat(prompt)
```

### 1.2 - Run the prompt

- Create a connection to the Azure OpenAI models
  - choose "gpt-5-mini"
  - Set the system prompt to "be terse"
- Use your prompt variable in the chat and save the output to `answer`
```{r}
conn <- chat_azure_openai(model = "gpt-5-mini", system_prompt = "be terse")
answer <- conn$chat(prompt)
```

### 1.3 - Modify a prompt with `glue`

The `glue` package allows you to put placeholders inside a string which will be
replaced with the content of a variable when glued together

**Example**
```{r}
animal1 <- "mouse"
animal2 <- "tiger"
animal3 <- "possum"
prompt <- glue("Is a {animal1} closer related to a {animal2} or a {animal3}")
cat(prompt)
```

Let's now use this in our prompt files ...

- Create a new file called `prompt_glue.txt`
- Write a simple prompt that has at least one `glue` placeholder
- Make sure to assign a variable with the same name and a relevant value
- Read in the file, paste all lines together and replace the placeholders
- Print your result to the console (or test it with the LLM)

```{r}
# Replace with your own file if you like
prompt_file <- system.file("rmarkdown", "templates", "edai_llm_extra", "skeleton", "prompt_glue.txt", package = "edAI")

disease <- "chickenpox"
prompt <- readLines(prompt_file) |> paste(collapse = "\n") |> 
  glue() |> cat()
```

There are cases when the text itself might contain `{` or `}` which should not
be interpreted by `glue` as a placeholder. An example is when you specify a JSON
template in your prompt. You can escape these by doubling the curly brackets and
`glue` will not interpret them as a variable and return a single brace.

In the prompt below, do the following

- Replace the hard-coded US state name with a variable (which you then create)
- Escape any curly brackets that should not be interpreted
- Use `glue` and `cat` to paste and print the result

```{r}
prompt <- paste(
  "How many people live in the state of {state} in the USA?",
  "If the state is not valid, return 0 for the population",
  "Your answer should be formatted as JSON that can be parsed directly using",
  'the following template: {{"validState": 1, "population": 1000}}'
  ) |> glue() |> cat()
```


## 2 - Using tools

There are certain tasks that LLMs are inherently bad at, for example when very 
computational or not easily really captured by language. However, if the input
for this other task can be described and the output is text, the LLM can run 
this task if needed using an R function or other tool registered by the user.

Because this is more advanced, we'll start with a full demo.

### 2.1 - Define an R function

The example function below will search PubMed for an article abstract given
a specific PubMed ID using the `rentrez` and `xml` libraries
```{r}
pubmed_abstract <- function(PMID) {
  abstract <- tryCatch({
    abstract <- entrez_fetch("pubmed", PMID, rettype = "xml")
    read_xml(abstract) |> xml_find_first(".//Abstract") |> xml_text()
  }, error = function(e){
    e
  })
  
  return(abstract)
}

```

### 2.2 - Test the function

- Navigate to PudMed an search for an article
- Use the PudMed ID (i.e. PMID) as input for the function above
- Check that the output matches what you see online

```{r}
pubmed_abstract("41071550")
```


### 2.3 - Convert the function into a tool for the LLM
Use the `tool` function from the `ellmer` package to create a new tool

- Check the `tool` documentation for `fun`, `description` and `arguments`

```{r}
pubmed_abstract_tool <- tool(
  pubmed_abstract,
  description = "Returns the abstract for a given Pubmed ID (PMID) or an error",
  arguments = list(
    PMID = type_string("The Pubmed ID (PMID) to get the abstract for")
  )
)
```
For a full list of types, check out https://ellmer.tidyverse.org/articles/tool-calling.html#tool-inputs-and-outputs

### 2.4 - Register the tool for use

- Create a new Azure connection object, set the model to gpt-5-mini and the 
  system prompt to be terse in its response
- Use the `register_tool` function from the `ellmer` package to register the
  tool you just created
- Now use the chat function to ask about the main conclusion of the pubmed 
  article with ID 40565474

```{r}
conn <- chat_azure_openai(model = "gpt-5-mini", system_prompt = "be terse")
conn$register_tool(pubmed_abstract_tool)

conn$chat("What is the main conclusion of the pubmed article with ID 40565474")
```

### 2.5 Try it yourself

Your aim is to have the LLM answer a question like this:
"How many Setosa species in the iris dataset have petal widths between 0.5 and 2"

Start by exploring the built-in iris dataset yourself first
```{r}

print(iris)

```


You are given the following function to get you started:
```{r}
iris_filter <- function(column, min_val = 0, max_val = 10){
  to_keep <- iris[,column] >= min_val & iris[,column] <= max_val
  iris[to_keep,"Species"] |> table() |> capture.output() |> paste(collapse = "\n")
}
```

TASK
- Explore what the function does and how it works (feel free to use AI if needed)
- Create a tool from this function that the LLM can use
- Register the tool and check if the LLM can answer the question

```{r}
iris_filter_tool <- tool(
  iris_filter,
  description = "Filter the iris data frame based on a range of column values and return the total number of rows that are retrained per species.",
  arguments = list(
    column = type_string("Column to filter on. Options are: Sepal.Length, Sepal.Width, Petal.Length, Petal.Width"),
    min_val = type_number("Minimum value to keep; default = 0"),
    max_val = type_number("Maximum value to keep; default = 10")
  )
)

conn <- chat_azure_openai(model = "gpt-5-mini", system_prompt = "be terse")
conn$register_tool(iris_filter_tool)

conn$chat("How many Setosa species in the iris dataset have petal widths between 0.5 and 2")
```
